A Horrocks-type theorem for even orthogonal $K_2$

### 2020-04-14 dl-paper.tex
Many fixes

Recall that to every irreducible root system $\Phi$ and a commutative ring $R$ one can associate two groups: the {\it simply-connected Chevalley group} $\GG(\Phi, R)$ (see e.\,g.~\cite[\S~3]{St71} or~\cite{VP}) and the {\it Steinberg group} $\St(\Phi, R)$ (see~\cref{sec:Steinberg-intro} for the definition). There is a well-defined map $\pi \colon \St(\Phi, R) \to \GG(\Phi, R)$ sending each generator $x_\alpha(\xi)$ to the elementary root unipotent $t_\alpha(\xi)$. The cokernel and the kernel of this map are, by definition, the $\K_1(\Phi, R)$ and $\K_2(\Phi, R)$-functors modeled on the Chevalley group $\GG(\Phi, -)$, see~\cite{St78}.

@@

Our main result is, thus, the following theorem, which is the orthogonal analogue of~\cite[Theorem~5.1]{Tu83} and the $\K_2$-analogue of~\cite[Theorem~6.8]{Su82} (cf. also with~\cite[Theorem~VI.5.2]{Lam10} and~\cite[Theorem~1.1]{St-poly}).
\begin{theorem}[Horrocks theorem for orthogonal $\K_2$]\label{thm:main} Let $A$ be a commutative ring in which $2$ is invertible. Then for any $\ell \geq 7$ the following commutative square is a pullback square in which all maps are injective:
\begin{equation} \nonumber
  \xymatrix{ \KO_2(2\ell, A) \ar[r] \ar[d] & \KO_2(2\ell, A[X]) \ar[d] \\ \KO_2(2\ell, A[X\inv]) \ar[r] & \KO_2(2\ell, A[X, X\inv]).}  
\end{equation} 
Moreover, the same assertion holds if one replaces the functor $\KO_2(2\ell, -)$ with $\K_2(\rD_\ell, -)$ or $\St(\rD_\ell, -)$. \end{theorem}
In the above statement $\KO_2(2\ell, -)$ denotes the unstable orthogonal $\K_2$-functor (see~\cref{sec:quillen}). Notice also that in the special case when $A=k$ is a field the assertion of the above theorem follows from the results of U.~Rehmann, J.~Hurrelbrink and J.~Morita (see~\cref{field-injectivity} below).

The proof of~\cref{thm:main} goes as follows. We notice that it suffices to prove the $\St(\rD_\ell, -)$-variant of the theorem. Moreover, the proof of the injectivity of
$j_+ \colon \St(\rD_\ell, A[X]) \to \St(\rD_\ell, A[X, X\inv])$ turns out to be the hardest part. After invoking the local-global principle~\cite[Theorem~2]{LS17} the proof reduces to the special case when $A$ is local.
Now if $M$ is the maximal ideal of $A$, the proof of the injectivity of $j_+$ comes down to proving injectivity of the following two maps:
\[ \xymatrix{ \St(\rD_\ell, A[X^{-1}]) \ar[r] & \St(\rD_\ell, A[X\inv] + M[X]) \ar[r] & \St(\rD_\ell, A[X, X\inv]). }\]
The injectivity of the second map is obtained in~\cref{thm41}. This is the only place in our proof which invokes the assumption that $2$ is invertible. The proof of~\cref{thm41} depends on the stability theorem for higher orthogonal K-groups and also on certain basic calculations with Grothendieck--Witt groups (these groups include the stable orthogonal K-groups as a special case, see~\eqref{GW-concrete}). Finally, the injectivity of the first map is obtained in~\cref{thm:P1glueing}, which is a direct generalization of~\cite[Proposition~4.3]{Tu83}. This part of the proof is obtained in somewhat greater generality and is applicable to all simply-laced root systems $\Phi$ containing a subsystem of type $\rA_4$.
%$\rD_\ell$, $\ell=5,6$ and also $\rE_\ell$, $\ell=6,7,8$.

@@

\begin{lemma} \label{field-injectivity} For $\Phi\neq\rG_2$ the map $\St(\Phi, k[X]) \to \St(\Phi, k[X, X^{-1}])$ is injective. Moreover, the intersection of the images of $\St(\Phi, k[X])$ and $\St(\Phi, k[X\inv])$ inside $\St(\Phi, k[X, X\inv])$ coincides with the image of $\St(\Phi, k)$. \end{lemma}
\begin{proof} The first assertion follows from the commutative diagram
\[\xymatrix{ & \K_2(\Phi, k) \ar[dl]_{\cong} \ar@{^{(}->}[dr] & \\
               \K_2(\Phi, k[X]) \ar[rr] &               & \K_2(\Phi, k[X, X^{-1}])} \]
and the injectivity of its diagonal arrows. Indeed, the left arrow is an isomorphism by the Korollar of~\cite[Satz~1]{Re75} and the right arrow is split injective by~\cref{K2-laurent-field}.

@@

Recall that the category of (commutative) pairs $\catname{Pairs}$ is defined as follows. Its objects are pairs $(R, I)$, in which $R$ is a commutative ring with $1$ and $I$ is a (not neccessarily proper) ideal of $R$. A morphism of pairs $f \colon (R, I) \to (R', I')$ is, by definition, a unit-preserving ring map $f \colon R \to R'$ such that $f(I) \subseteq I'$. Notice that the mapping $(R, I) \mapsto (R \to R/I)$ defines a functor from $\catname{Pairs}$ to the morphism category $\catname{CRings}^\rightarrow$.
If $(R, I)$ is such that $R$ is a local ring with maximal ideal $I$, we call such pair a {\it local pair}.

@@

Let $S \colon \catname{CRings} \to \catname{Groups}$ be a functor. Set $G_i = \Ker(S(p_i))$ and define {\it Loday's relativization} $S(R, I)$ as $ G_0 / [G_0, G_1]$. The map $S(p_1)$ induces a natural transformation $S(R, I) \to S(R)$. We denote this map by $\mu = \mu_{R,I}$ and its kernel by $C_S(R, I)$: \begin{equation} \label{LodayRelativization} \xymatrix{ 1 \ar[r] & C_S(R, I) \ar[r] & S(R, I) \ar^{\mu}[r] & S(R) \ar[r] & S(R/I) \ar[r] & 1.} \end{equation}

@@

For the rest of this subsection $S \xrightarrow{\pi} P \subseteq G$ is a triple of group-valued functors on the category of commutative rings satisfying the following assumptions:
\begin{enumerate} [label=\normalfont(A\arabic*)]
 \item \label{req:left-exact} $G(D_{R, I}) \cong G(R) \times_{G(R/I)} G(R)$.
 \item \label{req:coeq} For every pair $(R, I)$ the coequalizer of $S(p_0), S(p_1)$ is precisely $S(R) \to S(R/I)$.
 \item \label{req:subfunc} $P(R)$ is a perfect normal subgroup of $G(R)$.
 \item \label{req:uce} The map $ \pi_R \colon S(R) \to P(R)$ is a universal central extension for all $R$. In particular, $\HH_1(S(R), \ZZ) = \HH_2(S(R), \ZZ) = 0$.
\end{enumerate}

\begin{lemma}\label{lem:relativeH3}
 For every pair $(R, I)$ the map $\mu \colon S(R, I) \to S(R)$ is a universal relative central extension of $\nu \colon S(R) \to S(R/I)$. The group $C_S(R, I)$ is naturally isomorphic to the relative homology group $\HH_3(S(R/I), S(R); \ZZ)$.
\end{lemma}
\begin{proof}
The action of $S(R)$ on $S(D_{R, I})$ given by ${}^g h = S(\Delta)(g) \cdot h \cdot S(\Delta)(g)^{-1}$ induces an action of $S(R)$ on $S(R, I)$.
The map $\mu \colon S(R, I) \to S(R)$ from~\eqref{LodayRelativization} is an $S(R)$-map with respect to this action.
From~\ref{req:left-exact} and $\pi(G_i) \subseteq \Ker(G(p_i))$ we obtain that 
$G_0 \cap G_1 \subseteq \Ker(\pi_{D_{R, I}})$ hence it is a central subgroup of $S(D_{R,I})$ by~\ref{req:uce}. Thus, we have verified the assumptions of~\cite[Proposition~6]{Lo78} which asserts that the map $\mu$ is a universal relative central extension of the coequalizer $\nu = \mathrm{coeq}(d_0, d_1)$. Since $\nu$ coincides with $S(R) \to S(R/I)$ by~\ref{req:coeq}, we have completed the proof of the first assertion of the lemma.

Set $N = S(R)$, $Q = S(R/I)$, $C = \HH_3(Q, N; \ZZ)$. Recall from the proof of~\cite[Th{\'e}or{\`e}me~2]{Lo78} that to every relative central extension $(M, \mu)$ of $\nu$ with kernel $L$ one can associate a map of abelian groups $C \to L$. This map is obtained from the characteristic class $\xi(M, \mu)$ via the isomorphism $\HH^3(Q, N; L) \cong \mathrm{Hom}(C, L)$ of the universal coefficients theorem for cohomology (for this isomorphism we need the vanishing of $\HH_2(Q, N; \ZZ)$, which is a consequence of~\ref{req:uce}).

In the special case $M = S(R, I)$ this construction produces a map $C \to C_{S}(R, I)$ whose naturality in $(R, I)$ follows from~\cite[Proposition~3]{Lo78}. 
This map is an isomorphism by~\cite[Th{\'e}or{\`e}me~2]{Lo78}. \end{proof}

@@

For the rest of this section let us fix the following notation.
Let $A$ be an arbitrary commutative local ring with maximal ideal $M$ and residue field $k$.
Denote by $B = B_{A, M}$ the subring $A[X^{-1}] + M[X]$ of the ring $R = A[X, X^{-1}]$ and
by $I$ the ideal $M[X, X^{-1}]$ of $B$ (it is clear that $I$ is also an ideal of $R$).

\begin{lemma} \label{lem:prop41}
Assume additionally that the residue field $k$ is of characteristic $\neq 2$.
Then the canonical map $f\colon C(\rD_\ell, B, I) \to C(\rD_\ell, R, I)$ is surjective for $\ell \geq 7$. \end{lemma}
\begin{proof}
Writing the starting portion of the homology long exact sequence for the map $\St(\rD_\ell, R) \to \St(\rD_\ell, R/I)$ 
and using the isomorphisms of~\cref{lem:relativeH3} and~\cref{lem:lowerKgroups} we obtain the following commutative diagram:
\begin{equation*}\xymatrix{
 \KO_3(2\ell, B) \ar[r] \ar[d] & \KO_3(2\ell, k[X]) \ar[d]_{f'} \ar@{->>}[r] & \ar[d]_{f} C(\rD_\ell, B, I) \\
 \KO_3(2\ell, R) \ar[r]        & \KO_3(2\ell, k[X, X^{-1}]) \ar@{->>}[r]        & C(\rD_\ell, R, I).}\end{equation*}
By~\cref{Panin-stability} the map $f'$ can be identified with the canonical map $\GW_3^{[0]}(k[X]) \to \GW_3^{[0]}(k[X, X^{-1}])$.
By~\cref{bass-ft} $\GW_3^{[0]}(k[X, X^{-1}]) \cong \GW_3^{[0]}(k) \oplus \GW_2^{[3]}(k),$
but since the group $\GW_2^{[3]}(k)$ is trivial by~\cite[Lemma~2.2]{FRS12}, the map $f'$ (and hence $f$) is surjective.
\end{proof}

We will also need the following property of relative Steinberg groups which is a special case of a more general property discussed in~\cite[\S~2]{LS17}.
\begin{lemma}\label{lem:lemma32} Let $\Phi$ be a simply-laced root system of rank $\geq 3$,
Consider the following commutative square of canonical maps.
\[ \xymatrix{
    \St(\Phi, B, I) \ar[r] \ar[d] & \St(\Phi, B) \ar[d] \\
    \St(\Phi, R, I) \ar[r] \ar@{-->}^t[ur] & \St(\Phi, R) } \]
Then there exists a diagonal map $t$ which makes the diagram commute.   
\end{lemma} 
\begin{proof}
 Notice that $R$ is isomorphic to the principal localisation of $B$ at $X$
  and that $I$ is uniquely $X$-divisible in the sense of~\cite[\S~4]{LS17}.
 Thus, in the special case $\Phi = \rA_3$ the assertion of the lemma follows from~\cite[Theorem~3]{LS17}.
 In the general case the assertion of the lemma is a corollary of amalgamation theorem~\cite[Theorem~9]{S15}.
\end{proof}

\begin{externaltheorem} \label{thm41} Suppose that $2 \in A^\times$. Then for $\ell \geq 7$ the canonical map $\St(\rD_\ell, B) \to \St(\rD_\ell, R)$ is injective. \end{externaltheorem}
\begin{proof}
 Consider the following commutative diagram with exact rows obtained from~\eqref{LodayRelativization}:
\begin{equation*} \xymatrix{
 C(\rD_\ell, B, I) \ar[r]^{\lambda_B} \ar@{->>}[d]_{f} & \St(\rD_\ell, B, I) \ar[r]^{\mu_B} \ar[d]_{g} &
 \St(\rD_\ell, B) \ar[r]^{\nu_B} \ar[d]_{h} & \St(\rD_\ell, k[X]) \ar[d]_{i} \\
 C(\rD_\ell, R, I) \ar[r]^{\lambda_R}         & \St(\rD_\ell, R, I) \ar[r]^{\mu_R} \ar@{-->}[ur]^{t}&
 \St(\rD_\ell, R) \ar[r]^-{\nu_R}        & \St(\rD_\ell, k[X, X^{-1}]).
}\end{equation*}
The lifting $t$ in the above diagram is obtained from~\cref{lem:lemma32}.
Let $a$ be an element of $\Ker(h)$. Since $i$ is injective by~\cref{field-injectivity}, the element $a$
 also lies in $\Ker(\nu_B)$ and hence comes from some $b \in \St(\rD_\ell, B, I)$ via $\mu_B$.
Since $g(b) \in \Ker(\mu_R)$ there exists some $c \in C(\rD_\ell, R, I)$ such that $\lambda_R(c) = g(b)$. 
By~\cref{lem:prop41}, $f$ is surjective, therefore $c = f(d)$ for some $d \in C(\rD_\ell, R, I)$.
The required assertion now follows from the following calculation:
 \[ 1 = \mu_B\lambda_B(d) = tg\lambda_B(d) = t\lambda_Rf(d) =t(g(b)) = \mu_B(b) = a. \qedhere \]
\end{proof}

@@

We denote by $\overline{\St}(\Phi, R, I)$ the kernel of the map $\St(\Phi, R) \to \St(\Phi, R/I)$.
This group coincides with the image in $\St(\Phi, R)$ of the relative group $\St(\Phi, R, I)$ defined in~\cref{sec:quillen}.

\subsection{Generators of relative Steinberg groups}
Denote by $\St(\Phi, I)$ the subgroup of $\St(\Phi, R)$ generated as a group by root unipotents of level $I$.
It is clear that $\overline{\St}(\Phi, R, I)$ contains $\St(\Phi, I)$ and, in fact, is its normal closure.
We also denote by $\myol{H}(\Phi, R, I)$ the subgroup of $\overline{\St}(\Phi, R, I)$ generated by the semisimple root elements $h_\alpha(u)$ and symbols $\{u, v\}$, $u \in (1+I)^\times$, $v \in R^\times$, $\alpha\in \Phi$.

@@

It follows from~\cref{Kdecomp1} below that $K(A[X], M[X])$ coincides with the double commutator subgroup $[\overline{\St}(\Phi, A[X], M[X]), \overline{\St}(\Phi, A[X], XA[X])].$ Thus, if we replace relative Steinberg groups in the statement of~\cref{Kgen} with relative elementary groups, the resulting assertion turns into a special case of a much more general recent result of N.~Vavilov and Z.~Zhang (cf.~\cite[Theorem~1]{VZ18}).

@@

Throughout this section we use the following notations and conventions:
\begin{itemize}
 \item $A$ denotes an arbitarary commutative ring and $M$ is an ideal of $A$. Starting from subsection~\ref{sec:Pa0-basic} we also assume that $A$ is local and $M$ is the maximal ideal of $A$.
 \item We denote by $R$ the Laurent polynomial ring $A[X, X^{-1}]$. We set $t = X^{-1}$, so that $R = A[X, X^{-1}] = A[t, t^{-1}]$.
 \item $B = B_{A, M}$ denotes the subring $A[t] + M[t^{-1}] = A[X^{-1}] + M[X]$ of $R$.
 \item $I$ denotes the ideal $M[X, X^{-1}]$ of $R$. Notice that $I$ is also an ideal of $B$.
 \end{itemize}
\subsection{Presentation of Steinberg groups by homogeneous generators}
\label{sec:presentation}
We consider $R = A[t, t\inv]$ as a $\mathbb{Z}$-graded ring in which $t$ has degree $1$. This grading induces the grading on the subring $B \subseteq R$. For an integer $d$ we denote by $R_d$ (resp. $B_d$) the degree $d$ part of the ring $R$ (resp. $B$). Obviously, $B_d=M \cdot t^d$ for $d<0$, and $B_d=A \cdot t^d$ for $d\geq0$. With this notation, $B$ decomposes as $\oplus_{d\in\mathbb Z}B_d$ as an $A$-module.

@@

For a pair $(R, I)$ denote by $\E(\Phi, R, I)$ the {\it relative elementary subgroup} of $\GG(\Phi, R)$, i.\,e.
the image of the relative Steinberg group $\overline{\St}(\Phi, R, I)$ under the map $\pi \colon \St(\Phi, R) \to \GG(\Phi, R)$.
From the second property of the above proposition we immediately obtain the following group factorization.
\begin{corollary} For $\Phi$ as in the statement of~\cref{lemma33} one has
\[\E(\Phi, A[X\inv] + M[X]) = \E(\Phi, A[X], M[X]) \cdot \E(\Phi, A[X\inv]).\] \end{corollary}

>>>>

Recall that to every irreducible root system $\Phi$ and a commutative ring $R$ one can associate two groups: the {\it simply-connected Chevalley group} $\GG(\Phi, R)$ (see e.\,g.~\cite[\S~3]{St71} or~\cite{VP}) and the {\it Steinberg group} $\St(\Phi, R)$ (see~\cref{sec:Steinberg-intro} for the definition). There is a well-defined homomorphism $\pi \colon \St(\Phi, R) \to \GG(\Phi, R)$ sending each generator $x_\alpha(\xi)$ to the elementary root unipotent $t_\alpha(\xi)$. The cokernel and the kernel of this homomorphism (denoted $\K_1(\Phi, R)$ and $\K_2(\Phi, R)$, respectively) are, by definition, the values of the so-called $\mathrm{K}_1$ and $\mathrm{K_2}$-functors modeled on the Chevalley group $\GG(\Phi, -)$, see~\cite{St78}.

@@

Our main result is, thus, the following theorem, which is the orthogonal analogue of~\cite[Theorem~5.1]{Tu83} and the $\K_2$-analogue of~\cite[Theorem~6.8]{Su82} (cf. also with~\cite[Theorem~VI.5.2]{Lam10} and~\cite[Theorem~1.1]{St-poly}).
\begin{theorem}[Horrocks theorem for orthogonal $\K_2$]\label{thm:main} Let $A$ be a commutative ring in which $2$ is invertible. Then for any $\ell \geq 7$ the following commutative square is a pullback square in which all homomorphisms are injective:
\begin{equation} \nonumber
  \xymatrix{ \KO_2(2\ell, A) \ar[r] \ar[d] & \KO_2(2\ell, A[X]) \ar[d] \\ \KO_2(2\ell, A[X\inv]) \ar[r] & \KO_2(2\ell, A[X, X\inv]).}  
\end{equation} 
Moreover, the same assertion holds if one replaces the functor $\KO_2(2\ell, -)$ with $\K_2(\rD_\ell, -)$ or $\St(\rD_\ell, -)$. \end{theorem}
In the above statement $\KO_2(2\ell, -)$ denotes the unstable orthogonal $\K_2$-functor (see~\cref{sec:quillen}). Notice also that in the special case when $A=k$ is a field the assertion of the above theorem follows from the results of U.~Rehmann, J.~Hurrelbrink and J.~Morita (see~\cref{field-injectivity} below).

The proof of~\cref{thm:main} goes as follows. We notice that it suffices to prove the $\St(\rD_\ell, -)$-variant of the theorem. Moreover, the proof of the injectivity of
$j_+ \colon \St(\rD_\ell, A[X]) \to \St(\rD_\ell, A[X, X\inv])$ turns out to be the hardest part. After invoking the local-global principle~\cite[Theorem~2]{LS17} the proof reduces to the special case when $A$ is local.
Now if $M$ is the maximal ideal of $A$, the proof of the injectivity of $j_+$ comes down to proving injectivity of the following two homomorphisms:
\[ \xymatrix{ \St(\rD_\ell, A[X^{-1}]) \ar[r] & \St(\rD_\ell, A[X\inv] + M[X]) \ar[r] & \St(\rD_\ell, A[X, X\inv]). }\]
The injectivity of the second homomorphism is obtained in~\cref{thm41}. This is the only place in our proof which invokes the assumption that $2$ is invertible. The proof of~\cref{thm41} depends on the stability theorem for higher orthogonal K-groups and also on certain basic calculations with Grothendieck--Witt groups (these groups include the stable orthogonal K-groups as a special case, see~\eqref{GW-concrete}). Finally, the injectivity of the first homomorphism is obtained in~\cref{thm:P1glueing}, which is a direct generalization of~\cite[Proposition~4.3]{Tu83}. This part of the proof is obtained in somewhat greater generality and is applicable to all simply-laced root systems $\Phi$ containing a subsystem of type $\rA_4$.
%$\rD_\ell$, $\ell=5,6$ and also $\rE_\ell$, $\ell=6,7,8$.

@@

\begin{lemma} \label{field-injectivity} For $\Phi\neq\rG_2$ the homomorphism $\St(\Phi, k[X]) \to \St(\Phi, k[X, X^{-1}])$ is injective. Moreover, the intersection of the images of $\St(\Phi, k[X])$ and $\St(\Phi, k[X\inv])$ inside $\St(\Phi, k[X, X\inv])$ coincides with the image of $\St(\Phi, k)$. \end{lemma}
\begin{proof} The first assertion follows from the commutative diagram
\[\xymatrix{ & \K_2(\Phi, k) \ar[dl]_{\cong} \ar@{^{(}->}[dr] & \\
               \K_2(\Phi, k[X]) \ar[rr] &               & \K_2(\Phi, k[X, X^{-1}])} \]
and the injectivity of its diagonal arrows. Indeed, the left arrow is an isomorphism by the Korollar of~\cite[Satz~1]{Re75} and the right arrow is split injective by~\cref{K2-laurent-field}.

@@

Recall that the category of (commutative) pairs $\catname{Pairs}$ is defined as follows. Its objects are pairs $(R, I)$, in which $R$ is a commutative ring with $1$ and $I$ is a (not neccessarily proper) ideal of $R$. A morphism of pairs $f \colon (R, I) \to (R', I')$ is, by definition, a (unit-preserving) ring homomorphism $f \colon R \to R'$ such that $f(I) \subseteq I'$. Notice that the mapping $(R, I) \mapsto (R \to R/I)$ defines a functor from $\catname{Pairs}$ to the morphism category $\catname{CRings}^\rightarrow$.
If $(R, I)$ is such that $R$ is a local ring with maximal ideal $I$, we call such pair a {\it local pair}.

@@

Let $S \colon \catname{CRings} \to \catname{Groups}$ be a functor. Set $G_i = \Ker(S(p_i))$ and define {\it Loday's relativization} $S(R, I)$ as $ G_0 / [G_0, G_1]$. The homomorphism $S(p_1)$ induces a natural transformation $S(R, I) \to S(R)$. We denote this map by $\mu = \mu_{R,I}$ and its kernel by $C_S(R, I)$: \begin{equation} \label{LodayRelativization} \xymatrix{ 1 \ar[r] & C_S(R, I) \ar[r] & S(R, I) \ar^{\mu}[r] & S(R) \ar[r] & S(R/I) \ar[r] & 1.} \end{equation}

@@

For the rest of this subsection $S \xrightarrow{\pi} P \subseteq G$ is a triple of group-valued functors on the category of commutative rings satisfying the following assumptions:
\begin{enumerate} [label=\normalfont(A\arabic*)]
 \item \label{req:left-exact} $G(D_{R, I}) \cong G(R) \times_{G(R/I)} G(R)$.
 \item \label{req:coeq} For every pair $(R, I)$ the coequalizer of $S(p_0), S(p_1)$ is precisely $S(R) \to S(R/I)$.
 \item \label{req:subfunc} $P(R)$ is a perfect normal subgroup of $G(R)$.
 \item \label{req:uce} The homomorphism $ \pi_R \colon S(R) \to P(R)$ is a universal central extension for all $R$. In particular, $\HH_1(S(R), \ZZ) = \HH_2(S(R), \ZZ) = 0$.
\end{enumerate}

\begin{lemma}\label{lem:relativeH3}
 For every pair $(R, I)$ the homomorphism $\mu \colon S(R, I) \to S(R)$ is a universal relative central extension of $\nu \colon S(R) \to S(R/I)$. The group $C_S(R, I)$ is naturally isomorphic to the relative homology group $\HH_3(S(R/I), S(R); \ZZ)$.
\end{lemma}
\begin{proof}
The action of $S(R)$ on $S(D_{R, I})$ given by ${}^g h = S(\Delta)(g) \cdot h \cdot S(\Delta)(g)^{-1}$ induces an action of $S(R)$ on $S(R, I)$.
The homomorphism $\mu \colon S(R, I) \to S(R)$ from~\eqref{LodayRelativization} is an $S(R)$-homomorphism with respect to this action.
From~\ref{req:left-exact} and $\pi(G_i) \subseteq \Ker(G(p_i))$ we obtain that 
$G_0 \cap G_1 \subseteq \Ker(\pi_{D_{R, I}})$ hence it is a central subgroup of $S(D_{R,I})$ by~\ref{req:uce}. Thus, we have verified the assumptions of~\cite[Proposition~6]{Lo78} which asserts that the homomorphism $\mu$ is a universal relative central extension of the coequalizer $\nu = \mathrm{coeq}(d_0, d_1)$. Since $\nu$ coincides with $S(R) \to S(R/I)$ by~\ref{req:coeq}, we have completed the proof of the first assertion of the lemma.

Set $N = S(R)$, $Q = S(R/I)$, $C = \HH_3(Q, N; \ZZ)$. Recall from the proof of~\cite[Th{\'e}or{\`e}me~2]{Lo78} that to every relative central extension $(M, \mu)$ of $\nu$ with kernel $L$ one can associate a homomorphism of abelian groups $C \to L$. This homomorphism is obtained from the characteristic class $\xi(M, \mu)$ via the isomorphism $\HH^3(Q, N; L) \cong \mathrm{Hom}(C, L)$ of the universal coefficients theorem for cohomology (for this isomorphism we need the vanishing of $\HH_2(Q, N; \ZZ)$, which is a consequence of~\ref{req:uce}).

In the special case $M = S(R, I)$ this construction produces a homomorphism $C \to C_{S}(R, I)$ whose naturality in $(R, I)$ follows from~\cite[Proposition~3]{Lo78}. 
This homomorphism is an isomorphism by~\cite[Th{\'e}or{\`e}me~2]{Lo78}. \end{proof}

@@

For the rest of this section let us fix the following notation.
Let $A$ be an arbitrary commutative local ring with maximal ideal $M$ and residue field $k$.
Denote by $B$ the subring $A[X^{-1}] + M[X]$ of the ring $R = A[X, X^{-1}]$ and
by $I$ the ideal $M[X, X^{-1}]$ of $B$ (it is clear that $I$ is also an ideal of $R$).

\begin{lemma} \label{lem:prop41}
Assume additionally that the residue field $k$ is of characteristic $\neq 2$.
Then the canonical homomorphism $f\colon C(\rD_\ell, B, I) \to C(\rD_\ell, R, I)$ is surjective for $\ell \geq 7$. \end{lemma}
\begin{proof}
Writing the starting portion of the homology long exact sequence for the homomorphism $\St(\rD_\ell, R) \to \St(\rD_\ell, R/I)$ 
and using the isomorphisms of~\cref{lem:relativeH3} and~\cref{lem:lowerKgroups} we obtain the following commutative diagram:
\begin{equation*}\xymatrix{
 \KO_3(2\ell, B) \ar[r] \ar[d] & \KO_3(2\ell, k[X]) \ar[d]_{f'} \ar@{->>}[r] & \ar[d]_{f} C(\rD_\ell, B, I) \\
 \KO_3(2\ell, R) \ar[r]        & \KO_3(2\ell, k[X, X^{-1}]) \ar@{->>}[r]        & C(\rD_\ell, R, I).}\end{equation*}
By~\cref{Panin-stability} the homomorphism $f'$ can be identified with the canonical map $\GW_3^{[0]}(k[X]) \to \GW_3^{[0]}(k[X, X^{-1}])$.
By~\cref{bass-ft} $\GW_3^{[0]}(k[X, X^{-1}]) \cong \GW_3^{[0]}(k) \oplus \GW_2^{[3]}(k),$
but since the group $\GW_2^{[3]}(k)$ is trivial by~\cite[Lemma~2.2]{FRS12}, the homomorphism $f'$ (and hence $f$) is surjective.
\end{proof}

We will also need the following property of relative Steinberg groups which is a special case of a more general property discussed in~\cite[\S~2]{LS17}.
\begin{lemma}\label{lem:lemma32} Let $\Phi$ be a simply-laced root system of rank $\geq 3$,
Consider the following commutative square of canonical homomorphisms.
\[ \xymatrix{
    \St(\Phi, B, I) \ar[r] \ar[d] & \St(\Phi, B) \ar[d] \\
    \St(\Phi, R, I) \ar[r] \ar@{-->}^t[ur] & \St(\Phi, R) } \]
Then there exists a diagonal homomorphism $t$ which makes the diagram commute.   
\end{lemma} 
\begin{proof}
 Notice that $R$ is isomorphic to the principal localisation of $B$ at $X$
  and that $I$ is uniquely $X$-divisible in the sense of~\cite[\S~4]{LS17}.
 Thus, in the special case $\Phi = \rA_3$ the assertion of the lemma follows from~\cite[Theorem~3]{LS17}.
 In the general case the assertion of the lemma is a corollary of amalgamation theorem~\cite[Theorem~9]{S15}.
\end{proof}

\begin{externaltheorem} \label{thm41} Suppose that $2 \in A^\times$. Then for $\ell \geq 7$ the canonical homomorphism $\St(\rD_\ell, B) \to \St(\rD_\ell, R)$ is injective. \end{externaltheorem}
\begin{proof}
 Consider the following commutative diagram with exact rows obtained from~\eqref{LodayRelativization}:
\begin{equation*} \xymatrix{
 C(\rD_\ell, B, I) \ar[r]^{\lambda_B} \ar@{->>}[d]_{f} & \St(\rD_\ell, B, I) \ar[r]^{\mu_B} \ar[d]_{g} &
 \St(\rD_\ell, B) \ar[r]^{\nu_B} \ar[d]_{h} & \St(\rD_\ell, k[X]) \ar[d]_{i} \\
 C(\rD_\ell, R, I) \ar[r]^{\lambda_R}         & \St(\rD_\ell, R, I) \ar[r]^{\mu_R} \ar@{-->}[ur]^{t}&
 \St(\rD_\ell, R) \ar[r]^-{\nu_R}        & \St(\rD_\ell, k[X, X^{-1}]).
}\end{equation*}
The lifting $t$ in the above diagram is obtained from~\cref{lem:lemma32}.
Let $a$ be an element of $\Ker(h)$. Since $i$ is injective by~\cref{field-injectivity}, the element $a$
 also lies in $\Ker(\nu_B)$ and hence comes from some $b \in \St(\rD_\ell, B, I)$ via $\mu_B$.
Since $g(b) \in \Ker(\mu_R)$ there exists some $c \in C(\rD_\ell, R, I)$ such that $\lambda_R(c) = g(b)$. 
By~\cref{lem:prop41}, $f$ is surjective, therefore $c = f(d)$ for some $d \in C(\rD_\ell, R, I)$.
The required assertion now follows from the following calculation:
 \[ 1 = \mu_B\lambda_B(d) = tg\lambda_B(d) = t\lambda_Rf(d) =t(g(b)) = \mu_B(b) = a. \qedhere \]
\end{proof}

@@

We denote by $\overline{\St}(\Phi, R, I)$ the kernel of the homomorphism $\St(\Phi, R) \to \St(\Phi, R/I)$.
This group coincides with the image in $\St(\Phi, R)$ of the relative group $\St(\Phi, R, I)$ defined in~\cref{sec:quillen}.

\subsection{Generators of relative Steinberg groups}
Denote by $\St(\Phi, I)$ the subgroup of $\St(\Phi, R)$ generated as a group by all root subgroups $X_\alpha(I)$, $\alpha\in\Phi$.
It is clear that $\overline{\St}(\Phi, R, I)$ contains $\St(\Phi, I)$ and, in fact, is its normal closure.
We also denote by $\myol{H}(\Phi, R, I)$ the subgroup of $\overline{\St}(\Phi, R, I)$ generated by the semisimple root elements $h_\alpha(u)$ and symbols $\{u, v\}$, $u \in (1+I)^\times$, $v \in R^\times$, $\alpha\in \Phi$.

@@

It follows from~\cref{Kdecomp1} below that $K(A[X], M[X])$ coincides with the commutator subgroup $[\overline{\St}(\Phi, A[X], M[X]), \overline{\St}(\Phi, A[X], XA[X])].$ Thus, if we replace relative Steinberg groups in the statement of~\cref{Kgen} with relative elementary groups, the resulting assertion turns into a special case of a much more general recent result of N.~Vavilov and Z.~Zhang (cf.~\cite[Theorem~1]{VZ18}).

@@

Throughout this section we use the following notations and conventions:
\begin{itemize}
 \item $A$ denotes an arbitarary commutative ring and $M$ is an ideal of $A$. Starting from subsection~\ref{sec:Pa0-basic} we also assume that $A$ is local and $M$ is the maximal ideal of $A$.
 \item We denote by $R$ the Laurent polynomial ring $A[X, X^{-1}]$. We set $t = X^{-1}$, so that $R = A[X, X^{-1}] = A[t, t^{-1}]$.
 \item $B$ denotes the subring $A[t] + M[t^{-1}] = A[X^{-1}] + M[X]$ of $R$.
 \item $I$ denotes the ideal $M[X, X^{-1}]$ of $R$. Notice that $I$ is also an ideal of $B$.
 \end{itemize}
\subsection{Presentation of Steinberg groups by homogeneous generators}
\label{sec:presentation}
We consider $R = A[t, t\inv]$ as a $\mathbb{Z}$-graded ring in which $t$ has degree $1$. This grading induces the grading on the subring $B \subseteq R$. For an integer $d$ we denote by $R_d$ (resp. $B_d$) the degree $d$ part of the ring $R$ (resp. $B$). Obviously, $B_d=M \cdot t^d$ for $d<0$, and $B_d=A \cdot t^d$ for $d\geq0$. With this notation, $B$ decomposes as $\oplus_{d\in\mathbb Z}B_d$ as an $A$-module.

@@

For a pair $(R, I)$ denote by $\E(\Phi, R, I)$ the {\it relative elementary subgroup} of $\GG(\Phi, R)$, i.\,e.
the image of the relative Steinberg group $\overline{\St}(\Phi, R, I)$ under the homomorphism $\pi \colon \St(\Phi, R) \to \GG(\Phi, R)$.
From the second property of the above proposition we immediately obtain the following group factorization.
\begin{corollary} For $\Phi$ as in the statement of~\cref{lemma33} one has
\[\E(\Phi, A[X\inv] + M[X]) = \E(\Phi, A[X], M[X]) \cdot \E(\Phi, A[X\inv]).\] \end{corollary}
